---
title: "**MSBA 615 - R: Final Project**"
author: "Bradley Lowen"
date: "2024-12-10"
output: html_document
---

```{r,echo=FALSE, eval=TRUE,warning=FALSE,message=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(dslabs)
library(tibble)
library(gridExtra)
```

## **Importing and Cleaning Up NYC Airbnb Data**

**To begin, I will import the NYC Airbnb data from an Excel file.**

```{r,echo=TRUE, eval=TRUE,warning=FALSE,message=FALSE}
airbnb <- read_excel("C:/Users/bwlowe01/Desktop/MSBA/MSBA 615 - R for Analytics/AB_NYC_2019.xlsx")

```
**Then, looking at the data, I can omit some entire columns that do not provide
data that would help with analysis. The columns "id", "host_id", "name",
"host_name","latitude","longitude", "last_review", and "reviews_per_month" 
all do not meaningfully contribute to the analysis of this
data. The "id" and "host_id" columns are accounted for much more efficiently
in the "calculated_host_listings_count" column.
The latitude and longitude location data is redundant due to both the
"neighbourhood_group" and "neighbourhood" columns, and the "reviews_per_month"
and "last_review"columns provide data that is not pertinent enough in
light of the other categories.These would only be useful if you are attempting
to analyze review trends, or focus specifically on reviews.
The only caveat to removing "id" is that we should explore the
data for duplicates before removing this category. Some listings could
be identical in other ways just by chance.**

```{r,echo=TRUE, eval=TRUE,warning=FALSE,message=FALSE}
sum(duplicated(airbnb))

#no duplicates found, we can remove "id" and clean the data

airbnbclean <- airbnb |> select(-"id", -"host_id", -"name", -"host_name", -"latitude", -"longitude", -"last_review", -"reviews_per_month")

```
**Now, I can check the remaining data for NA values and outliers.
We will check for outliers using the IQR technique, and visualize with a box plot.**

```{r,echo=TRUE, eval=TRUE,warning=FALSE,message=FALSE}
navalues <- colSums(is.na(airbnbclean))
navalues

```
**There are no NA values.**


```{r,echo=TRUE, eval=TRUE,warning=FALSE,message=FALSE}
findoutliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE) 
  Q3 <- quantile(x, 0.75, na.rm = TRUE) 
  IQR <- Q3 - Q1                       
  lowerq <- Q1 - 1.5 * IQR        
  upperq <- Q3 + 1.5 * IQR        
  return(x < lowerq | x > upperq) 
}

airbnbclean$outlierprice <- findoutliers(airbnbclean$price)
airbnbclean$outlierminnights <- findoutliers(airbnbclean$minimum_nights)
airbnbclean$outlierreviews <- findoutliers(airbnbclean$number_of_reviews)
airbnbclean$outlieravailability <- findoutliers(airbnbclean$availability_365)

sum(airbnbclean$outlierprice)        
sum(airbnbclean$outlierminnights) 
sum(airbnbclean$outlierreviews)     
sum(airbnbclean$outlieravailability) 

#note that for price, IQR may not count 0 as an outlier, but for our
#purposes, 0 is not a useful price for analysis, so we will remove it. 

airbnbclean2 <- airbnbclean |>
  filter(price > 0)
```

## Visualizing Outliers

**Let's look at the outliers using box plots:**
```{r,echo=TRUE, eval=TRUE,warning=FALSE,message=FALSE}
boxplotprice <- ggplot(airbnbclean2, aes(y = price)) +
  geom_boxplot(fill = "red", width = 5) +
  labs(title = "Prices", y = "Price") + 
  theme_minimal()

boxplotminnights <- ggplot(airbnbclean2, aes(y = minimum_nights)) +
  geom_boxplot(fill = "lightblue", width = 5) +
  labs(title = "Min Nights", y = "Number of Minimum Nights") +
  theme_minimal()

boxplotreviews <- ggplot(airbnbclean2, aes(y = number_of_reviews)) +
  geom_boxplot(fill = "lightgreen", width = 5) +
  labs(title = "Reviews", y = "Number of Reviews") +
  theme_minimal()

boxplothost <- ggplot(airbnbclean2, aes(y = calculated_host_listings_count)) +
  geom_boxplot(fill = "brown", width = 5) +
  labs(title = "Host Listings", y = "Number of Listings") + 
  theme_minimal()

boxplotavail <- ggplot(airbnbclean2, aes(y = availability_365)) +
  geom_boxplot(fill = "orange", width = 5) +
  labs(title = "Availability", y = "Days of Availability") +
  theme_minimal()

grid.arrange(boxplotprice, boxplotminnights, boxplotreviews, boxplothost, boxplotavail, ncol = 5)

```

**Now to remove the outliers and temporary columns from the dataset.**

```{r,echo=TRUE, eval=TRUE,warning=FALSE,message=FALSE}
airbnbnooutliers <- airbnbclean2 |>
  filter(
    !outlierprice & 
    !outlierminnights & 
    !outlierreviews & 
    !outlieravailability)

airbnbnooutliers <- airbnbnooutliers |>
  select(-outlierprice, -outlierminnights, -outlierreviews, -outlieravailability)

```

# **Interpreting the Cleaned Data**

## Price

**Now, we can analyze the price distribution with a histogram and boxplot.**

```{r,echo=TRUE, eval=TRUE,warning=FALSE,message=FALSE}
ggplot(airbnbnooutliers, aes(x = price)) +
  geom_histogram(binwidth = 10, fill = "red", color = "black") +
  ggtitle("Price Distribution (Histogram)") +
  xlab("Price") +
  ylab("Count")
ggplot(airbnbnooutliers, aes(y = price)) +
  geom_boxplot(fill = "green") +
  ggtitle("Price Distribution (Boxplot)") +
  ylab("Price")
```


**From these two plots, we can see that a majority of listings are between
$50 and $150.**

## Neighborhood Groups
Let's look at the average price per neighborhood group and the percentage of
listings per neighborhood group.
```{r,echo=TRUE, eval=TRUE,warning=FALSE,message=FALSE}
avgneighborhood <- airbnbnooutliers |>
  group_by(neighbourhood_group) |>
  summarise(AvgPrice = mean(price))

ggplot(avgneighborhood, aes(x = reorder(neighbourhood_group, AvgPrice), y = AvgPrice)) +
  geom_bar(stat = "identity", fill = "orange") +
  ggtitle("Average Price by Neighborhood Group") +
  xlab("Neighborhood Group") +
  ylab("Average Price") +
  coord_flip()

listbyhood <- airbnbnooutliers |>
  count(neighbourhood_group) |>
  arrange(desc(n))


listbyhood <- listbyhood |>
  mutate(Percentage = round(n / sum(n) * 100, 1)) 

listbyhood$label <- paste0(listbyhood$neighbourhood_group, " (", listbyhood$Percentage, "%)")

ggplot(listbyhood, aes(x = "", y = n, fill = label)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +  
  theme_void() +  
  labs(
    title = "Percentage of Listings by Neighborhood Group",
    fill = "Neighborhood Group (with %)") +
  scale_fill_discrete(name = "Neighborhood Group")
```

**As we can see from the histogram, Manhattan has the highest average price, 
followed by, in order, Brooklyn, Queens, Staten Island, and the Bronx.
From our pie chart, we see that Brooklyn has the highest number of listings,
followed by, in order, Manhattan, Queens, the Bronx, and Staten Island.
Brooklyn and Manhattan comprise over 83% of all listings.**

## Neighborhoods
**To go into even finer detail, we can look at some of the top neighborhoods
for Airbnb listings.** 
```{r,echo=TRUE, eval=TRUE,warning=FALSE,message=FALSE}
topneighborhoods <- airbnbnooutliers |>
  count(neighbourhood) |>
  arrange(desc(n)) |>
  top_n(10, n)

ggplot(topneighborhoods, aes(x = reorder(neighbourhood, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Neighborhoods by Number of Listings",
    x = "Neighborhood",
    y = "Number of Listings")

topneighborhoodprices <- airbnbnooutliers |>
  group_by(neighbourhood) |>
  summarise(AvgPrice = mean(price)) |>
  arrange(desc(AvgPrice)) |>
  top_n(10, AvgPrice)

ggplot(topneighborhoodprices, aes(x = reorder(neighbourhood, AvgPrice), y = AvgPrice)) +
  geom_bar(stat = "identity", fill = "brown") +
  coord_flip() +
  labs(
    title = "Top 10 Neighborhoods by Average Price",
    x = "Neighborhood",
    y = "Average Price"
  ) +
  theme_minimal()
```

**As we can see comparing these two plots, there is no overlap whatsoever between
the top 10 neighborhoods for listings and the top 10 neighborhoods by average
price. The top 10 neighborhoods by listing count show us the areas with the
highest demand, while the average price top 10 helps us target more affluent
areas.** 

##  Room Type and Average Price

**Let's examine the average price of an Airbnb by room type.**
```{r,echo=TRUE, eval=TRUE,warning=FALSE,message=FALSE}

avgpriceroom <- airbnbnooutliers |>
  group_by(room_type) |>
  summarise(AvgPrice = mean(price))

print(avgpriceroom)

ggplot(avgpriceroom, aes(x = room_type, y = AvgPrice, fill = room_type)) +
  geom_bar(stat = "identity") +
  ggtitle("Average Price by Room Type") +
  xlab("Room Type") +
  ylab("Average Price") +
  theme_minimal()
```


**As we can see, an entire home/apartment is the most expensive option, by a wide
margin. Private rooms come in second, while shared rooms are cheapest.**

## Availability

**Let's examine room availability.**
```{r,echo=TRUE, eval=TRUE,warning=FALSE,message=FALSE}
ggplot(airbnbnooutliers, aes(x = availability_365)) +
  geom_histogram(binwidth = 20, fill = "purple", color = "black") +
  ggtitle("Availability Distribution") +
  xlab("Availability (Days per Year)") +
  ylab("Number of Listings")


```

**From our availability distribution, we can see that NYC has a high demand
for temporary housing, with over half of the listings already having 
no availability.**


# Predicting Price with Linear Regression

```{r,echo=TRUE, eval=TRUE,warning=FALSE,message=FALSE}
modeldata <- airbnbnooutliers |>
  select(price, number_of_reviews, availability_365, room_type, neighbourhood_group)

modeldata$room_type <- as.factor(modeldata$room_type)
modeldata$neighbourhood_group <- as.factor(modeldata$neighbourhood_group)

set.seed(123)
trainIndex <- sample(1:nrow(modeldata), 0.8 * nrow(modeldata))  
traindata <- modeldata[trainIndex, ]
testdata <- modeldata[-trainIndex, ]

lrmodel <- lm(price ~ number_of_reviews + availability_365 + room_type + neighbourhood_group, data = traindata)

predictions <- predict(lrmodel, newdata = testdata)

comparison <- data.frame(Actual = testdata$price, Predicted = predictions)
head(comparison)

ggplot(comparison, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.5, color = "lightblue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  ggtitle("Actual vs Predicted Prices") +
  xlab("Actual Price") +
  ylab("Predicted Price") +
  theme_minimal()

summary(lrmodel)

```


## Observations from the Model
**From our linear regression model, we can see that number of reviews, 
room type, and neighborhood all have a significant influence on price, with 
room type and neighborhood group having the largest impact on the price of an
Airbnb. From our model fit R-squared value (.4562), there can be improvement. 
We are ideally looking for over .7 for this value, so some factors could be
missing. These could include, for example, proximity to public transportation
stations, square footage, or proximity to restaurants or tourist destinations.**

```{r}

```